---
title: 'Coding Challenge 6: Loops and Functions'
author: "Katie Temple"
date: "2025-03-27"
output: 
  word_document:
    toc: true 
  md_document:
      variant: gfm
---

## Question 1
2 pts. Regarding reproducibility, what is the main point of writing your own functions 
and iterations? 

### Answer 1

The main point in writing your own functions is to reduce potential error
through typos, copy and paste mistakes, , missremembering a formula, and that a lot 
of it is easier than doing by hand. Additionally, it clearly lays out why you're 
doing what you're doing and those trying to reproduce your analysis should be able 
to plug in the same values using the function. 
The same can be said for when needing have/write iterative values of something. 


## Question 2
2 pts. In your own words, describe how to write a function and a for loop in R 
and how they work. Give me specifics like syntax, where to write code, and how 
the results are returned. 

### Answer 2
When writing your own function by hand, you need to first know what you want to 
use the function for. In the example in the homework, we wanted to convert Fahrenheit
temperatures to Celsius and used the standard conversion equation to do so. 

#### Functions Answer
```{r}
#Name your function and tell r you want to make a function.
# Example: converting ounces to grams: The equation to do that is ounces*28.35. 
#The first line of code is also where you want to name your input/what you want to manipulate 
#(in this case we want to convert ounces.)
ounces_to_grams <- function (ounces){  #name your function and then tell r you want to make a function with "
  #function() and inside the parenthesis put what you want to manipulate. Then use the squiggly 
  #brackets and put enter so that they encapsulate the next steps.
  grams <- (ounces*28.35) #write the equation by name with your variable that you're 
  #converting. In this case ounces. The output is name of the converted variable you want returned.
  return(grams) #here you're saying that you want the value of the converted variable to grams returned
} #close the function. This will show your Global Environment under the functions aspect.

#now use your function. Convert some ounces to grams
ounces_to_grams(16) #in this case you're converting 16 ounces to grams
#[1] 453.6
#Now check
16*28.35
#[1] 453.6
```

#### For loops Answer
When writing for loops, you need to be explicit. For loops will loop pack to (in this example)
each number in the sequence and manipulate it the way you wanted (in the example below multiplied by the ounces to gram conversion of 28.35). It will loop back until it has addressed each part of the sequence you've given. Each line generates a single row dataframe of the output unless you bind in (such as with rbind).
```{r}
for(i in 0:10) {#i is what you want to iterate. Then do squiggle bracket to start and enter what you want to do
  print (i*28.35) #have i and say what you want to happen to i. In this case i want to multiply it by 28.35
  #the print command is used to so that you can see the output. You have to be explicit with loops such as saying you want the output to be displayed. This is where you put your function (if you want your i and product of i in regard to function)
}
#[1] 0
#[1] 28.35
#[1] 56.7
#[1] 85.05
#[1] 113.4
#[1] 141.75
#[1] 170.1
#[1] 198.45
#[1] 226.8
#[1] 255.15
#[1] 283.5

#Within the squiggle brackets, you can also do things like create dataframes. A way to think about it 
#is each line should be what you want to happen and then the next is what you want to happen next in regard to the previous line. 
```


## Question 3
2 pts. Read in the Cities.csv file from Canvas using a relative file path.
```{r}
cities <- read.csv("Cities.csv", na.strings = "na")
str(cities)
```

## Question 4
6 pts. Write a function to calculate the distance between two pairs of coordinates 
based on the Haversine formula (see below). The input into the function should be
lat1, lon1, lat2, and lon2. The function should return the object distance_km. 
All the code included needs to go into the function.

```{r}
#Write a function using the steps in the assignment.
# convert to radians
distance <-function (lat1, lon1, lat2, lon2){
# convert to radians
rad.lat1 <- lat1 * pi/180
rad.lon1 <- lon1 * pi/180
rad.lat2 <- lat2 * pi/180
rad.lon2 <- lon2 * pi/180

# Haversine formula
delta_lat <- rad.lat2 - rad.lat1
delta_lon <- rad.lon2 - rad.lon1
a <- sin(delta_lat / 2)^2 + cos(rad.lat1) * cos(rad.lat2) * sin(delta_lon / 2)^2
c <- 2 * asin(sqrt(a)) 

# Earth's radius in kilometers
earth_radius <- 6378137

# Calculate the distance
distance_km <- (earth_radius * c)/1000
return (distance_km)
}
```

## Question 5
5 pts. Using your function, compute the distance between Auburn, AL and New York City 

 a.	Subset/filter the Cities.csv data to include only the latitude and longitude values you need and    input as input to your function. 

 b.The output of your function should be 1367.854 km
```{r}
library(tidyverse)
str(cities)
View(cities)
latau=cities$lat[cities$city=="Auburn"]
lonau=cities$long[cities$city=="Auburn"]
latny=cities$lat[cities$city=="New York"]
lonny=cities$long[cities$city=="New York"]

distance(latau,lonau, latny, lonny)
#[1] 1367.854
```
## Question 6
6 pts. Now, use your function within a for loop to calculate the distance between all other cities in the data. The output of the first 9 iterations is shown below. 
```{r}
#need to change the function so that 
nm<-unique(cities$city)

distance2 <-function (lat1, lon1, lat2, lon2){
lat1=cities$lat[cities$city=="Auburn"]
lon1=cities$long[cities$city=="Auburn"]
lat2=cities$lat[cities$city==nm[[i]]]
lon2=cities$lon[cities$city== nm[[i]]]
  # convert to radians
rad.lat1 <- lat1 * pi/180
rad.lon1 <- lon1 * pi/180
rad.lat2 <- lat2 * pi/180
rad.lon2 <- lon2 * pi/180

# Haversine formula
delta_lat <- rad.lat2 - rad.lat1
delta_lon <- rad.lon2 - rad.lon1
a <- sin(delta_lat / 2)^2 + cos(rad.lat1) * cos(rad.lat2) * sin(delta_lon / 2)^2
c <- 2 * asin(sqrt(a)) 

# Earth's radius in kilometers
earth_radius <- 6378137

# Calculate the distance
distance_km <- (earth_radius * c)/1000
return (distance_km)
}

distances.df <-NULL
for(i in seq_along(nm)){
  result_i<- data.frame(distance2())
  distances.df<-rbind.data.frame(distances.df, result_i)
}
print(distances.df)
```

##Question 7
link:
[link to GitHub](https://github.com/temkat/TempleReproducibilityClass2025/tree/5bdfb88dca4e0da5e664e870b14e8216a988b9af/LoopsFunctions/CodingChallenge6)