---
title: "Coding Challenge 5"
author: "Katie Temple"
date: "2025-03-20"
output: 
  html_document:
     toc: true 
  md_document:
     variant: gfm  
  word_document:
    toc: true
---


### Question 1
3pts. Download two .csv files from Canvas called DiversityData.csv and Metadata.csv, and read them into R using relative file paths.
```{r}
library(tidyverse)
diversity<- read.csv ("DiversityData.csv", na.strings = "na")
str(diversity)
meta<- read.csv ("Metadata.csv", na.strings= "na")
str(meta)
```

### Question 2
4 pts. Join the two dataframes together by the common column ‘Code’. Name the resulting dataframe alpha.
```{r}
alpha <-left_join(diversity, meta, by= "Code") #join the two dataframes into one and name it alpha.
head(alpha)
```
### Question 3
4 pts. Calculate Pielou’s evenness index: Pielou’s evenness is an ecological parameter calculated by the Shannon diversity index (column Shannon) divided by the log of the richness column.

   a.	Using mutate, create a new column to calculate Pielou’s evenness index. 
   b.	Name the resulting dataframe alpha_even.


```{r}
alpha_even<-alpha%>% #create new dataframe called alpha_even(b)
 mutate(even=shannon/log(richness)) #mutate to create new column to calculate 
 #Pieloe's evenness by dividing the shannon column by the log of the richness column (a).

head(alpha_even) #check data frame.
 

```
### Question 4
4. Pts. Using tidyverse language of functions and the pipe, use the summarise function and tell me the mean and standard error evenness grouped by crop over time.

 a.	Start with the alpha_even dataframe
 b.	Group the data: group the data by Crop and Time_Point.
 c.	Summarize the data: Calculate the mean, count, standard deviation, and standard error for     the even variable within each group.
 d.	Name the resulting dataframe alpha_average

```{r}
alpha_average<- alpha_even%>% #use the alpha dataframe and create new dataframe cammed alpha_average
  group_by(Crop, Time_Point)%>% #grouping by crop and time_point
  summarise(
    mean.even=mean(even), #mean of the Pielou’s evenness index from the new coumn named even
    n=n(),
    std.dev=sd(even), 
    std.err = (std.dev/sqrt(n))
  )
```

### Question 5
4. Pts. Calculate the difference between the soybean column, the soil column, and the difference between the cotton column and the soil column

  a.	Start with the alpha_average dataframe
  b.	Select relevant columns: select the columns Time_Point, Crop, and mean.even.
  c.	Reshape the data: Use the pivot_wider function to transform the data from long to wide        format, creating new columns for each Crop with values from mean.even.
  d.	Calculate differences: Create new columns named diff.cotton.even and diff.soybean.even        by calculating the difference between Soil and Cotton, and Soil and Soybean,
      respectively.
  e.	Name the resulting dataframe alpha_average2
```{r}
alpha_average2<- alpha_average%>% #new new data frame and use alpha_average as ref.
                 select(Time_Point, Crop, mean.even)%>% # select relevant columns
                 pivot_wider(names_from = Crop, values_from= mean.even)%>% #convert data to long form wide form
                 mutate(diff.cotton.even = Soil - Cotton)%>% #new column with difference from soil and cotton
                 mutate(diff.soybean.even = Soil - Soybean) #new column with difference between soil and soybean
                 
  
```

### Question 6
4 pts. Conecting it to plots.
 
 a.	Start with the alpha_average2 dataframe
 b.	Select relevant columns: select the columns Time_Point, diff.cotton.even, and
    diff.soybean.even
 c.	Reshape the data: Use the pivot_longer function to transform the data from wide to long       format, creating a new column named diff that contains the values from diff.cotton.even
    and diff.soybean.even.
    
   i.	This might be challenging, so I’ll give you a break. The code is below

 d. Create the plot: Use ggplot and geom_line() with ‘Time_Point’ on the x-axis, the column       ‘values’ on the y-axis, and different colors for each ‘diff’ category. The column name        ‘values’ come from the pivot_longer. The resulting plot should look like the one to the       right
```{r}
#still need to remove background!!!
alpha_average2%>% #part 6a
  select(Time_Point,diff.cotton.even, diff.soybean.even)%>% #6b
  pivot_longer(c(diff.cotton.even, diff.soybean.even), names_to = "diff")%>%#6c
  ggplot( aes(x = Time_Point, y = value, color = diff, group = diff))+
  geom_line()+
  xlab("Time (hrs)")+
  ylab ("Difference from soil in Pielou's evenness")
```

### Question 7
Commit and push a gfm .md file to GitHub inside a directory called Coding Challenge 5. 
Provide me a link to your github written as a clickable link in your .pdf or .docx

[Katie Temple's GitHub Repository Link]()

